\textbf{ Analytic Light Curves in Reflected Light: \\% Phase Curves, Occultations, and Non-Lambertian Scattering for Spherical Planets and Moons }

### 2021-03-31 tex/ms.tex
linenos, citations, misc

% !TeX root = ./ms.tex
\documentclass[modern]{aastex62}

@@

@@

Our goal in this paper is to derive analytic expressions for the
flux received by a distant observer from a sphere of non-uniform albedo
illuminated by a monochromatic source that may or may not be
occulted by a (possibly different) spherical body. This applies, for example,
to the case of planetary phase curves, secondary
eclipse (occultation) light curves, and moon-moon, planet-moon, and
planet-planet occultations (in the Solar System or not),
all seen in reflected light. We derive all expressions in
the limit that the reflecting
body is Lambertian, i.e., it scatters light isotropically, but we relax
this assumption in later sections.
We model the general case of an
intensity that varies across the surface of the body according to a
spatially-dependent
albedo $A$. Throughout this paper, we will take $A$ to mean the \emph{spherical}
albedo, the fraction of power incident on a body at a given wavelength
that is scattered back out to space (in all directions).
Note that the spherical albedo
is closely related to the Bond albedo: the Bond albedo is the stellar
flux-weighted integral of $A(\lambda)$ over all wavelengths $\lambda$
\citep[see, e.g.,][]{Seager2010}.

@@

In the limit that the angular size of the star as seen from the planet is
small, the illumination profile on the surface of the planet will
decrease as the cosine of the angle between the surface normal and the star,
reaching zero at the day/night terminator, an angle $\nicefrac{\pi}{2}$
away from the sub-stellar point. However, if the star is sufficiently large
and the planet is sufficiently close-in, rays originating from near the limb
of the star will reach points on the planet surface beyond this angle.
If the stellar radius $R_\star$
is larger than the planet radius $R_\mathrm{p}$, the angular extent of the true
day/night terminator past $\nicefrac{\pi}{2}$ is given by
%
\begin{proof}{tau}
    \label{eq:tau}
    \tau &= \arcsin\left( \frac{1 - \nicefrac{R_\mathrm{p}}{R_\star}}{\nicefrac{a}{R_\star}} \right)
\end{proof}
%
where $a$ the semi-major axis of the orbit (where we implictly assume
the eccentricity is zero).
For planets like the Earth, this quantity is only about $0.26^\circ$,
resulting in a negligible effect on the planet's light curve. However,
for very close-in planets, the effect can be significant. For instance,
the hot Jupiter Kelt-9b has $\nicefrac{R_\mathrm{p}}{R_\star} = 0.083$ and
$\nicefrac{a}{R_\star} = 3.16$
planet \citep{Wong2019}. Assuming zero eccentricity and ignoring any
stellar oblateness
\citep[see][]{Ahlers2020},
the day/night terminator extends $\tau \approx 17^\circ$ past the limb
of that planet. Figure~\ref{fig:extended} shows the illumination profile
of this planet in a Mollweide projection, with the sub-stellar point at
the center, for the point source approximation (left) and accounting for
the finite size of the star (right). In addition to the
displaced day/night terminator, the main difference between
the two profiles is the sub-stellar intensity, which is significantly higher
in the extended source case. This is due to the simple fact that in the point
source case the illumination source is placed at the center of the star,
which is one stellar radius farther from the planet than the point closest
to the planet (the sub-planetary point) in the extended source case.
Once accounting for this difference, the fractional change in the intensity
on the planet away from the sub-stellar point is similar in both cases,
and the intensity anywhere beyond $\nicefrac{\pi}{2}$ is less than one-tenth
the peak value.

@@

Finally, we would like to emphasize that while spherical harmonics are
a convenient basis for the purpose of computing light curves, they
have a significant drawback when it comes to modeling real planetary
surfaces: it can be difficult to strictly enforce physical
values of the albedo everywhere on the surface map when doing inference.
That is because there is no analytic way to determine whether a
spherical harmonic representation is positive-valued
(or restricted to a given range) everywhere on the sphere.
Instead, this must be checked numerically, by evaluating the function on a
discrete grid. This makes it somewhat cumbersome to implement positivity as a prior
when doing inference; in particular, this prior cannot be expressed as
a Gaussian, so the analytic expression for the posterior
discussed in \S\ref{sec:linearity} will generally have nonzero support for
negative albedo values. This is particularly problematic when the data is
not very constraining and the posterior is prior-dominated, as
positivity can be an \emph{extremely} informative prior
\citep[e.g.,][]{Fienup1978}.
We therefore recommend that in such cases
users of the \starry algorithm use HMC/MCMC to do inference, either
(1) sampling over the spherical
harmonic coefficients $\mathbf{y}$ and imposing a uniform prior in the range
$[0, 1]$ on the albedo values $\mathbf{a}$
evaluated on a discrete grid on the sphere or
(2) sampling over the albedo values $\mathbf{a}$ with the same uniform prior,
but using $\mathbf{y}$ to compute the actual light curve model
\citep{Bartolic2021}. In both
cases, there exists a linear operator that transforms between
$\mathbf{y}$ and $\mathbf{a}$:
%
\begin{align}
    \label{eq:Py}
    \mathbf{a} = \mathbf{P} \mathbf{y}
\end{align}
%
and
%
\begin{align}
    \label{eq:PInvy}
    \mathbf{y} = \mathbf{P}^+ \mathbf{a}
    \quad,
\end{align}
%
where
%
\begin{align}
    \label{eq:PInv}
    \mathbf{P}^+ =
    (\mathbf{P}^\top \mathbf{P} + \lambda \boldsymbol{\mathsf{I}})^{-1} \mathbf{P}^\top
\end{align}
%
is the pseudoinverse of $\mathbf{P}$, with (small) regularization parameter
$\lambda$ and where $\boldsymbol{\mathsf{I}}$ is the identity matrix.
%
Each row of the matrix $\mathbf{P}$ is constructed from the value
of each of the spherical harmonics at the corresponding point on the
grid; both $\mathbf{P}$ and its inverse may be pre-computed for
efficiency. In both cases, the grid should be fine enough to ensure
positivity over most of the sphere but not so fine as to throttle the
computation; as a rule of thumb, we find that grids with ${\sim}4$ times
as many pixels
as spherical harmonic coefficients are sufficient.
%
The \href{https://starry.readthedocs.io}{documentation}
includes tutorials on how to implement this in practice.

>>>>

% !TeX root = ./ms.tex
\ifdefined\LINENOS
    \documentclass[modern,linenumbers]{aastex62}
\else
    \documentclass[modern]{aastex62}
\fi

@@

% Force line numbers in paragraphs containing equations
\ifdefined\LINENOS
    \usepackage{lineno}
    \linenumbers
    \let\oldalign\align
    \let\oldendalign\endalign
    \renewenvironment{align}{%
        \linenomath%
            \oldalign%
    }{%
        \endlinenomath%
            \oldendalign%
    }
\fi

@@

Our goal in this paper is to derive analytic expressions for the
flux received by a distant observer from a sphere of non-uniform albedo
illuminated by a monochromatic source that may or may not be
occulted by a (possibly different) spherical body. This applies, for example,
to the case of planetary phase curves, secondary
eclipse (occultation) light curves, and moon-moon, planet-moon, and
planet-planet occultations \citep[in the Solar System or not; e.g.,][]{Cabrera2007,Luger2019b},
all seen in reflected light. We derive all expressions in
the limit that the reflecting
body is Lambertian, i.e., it scatters light isotropically, but we relax
this assumption in later sections.
We model the general case of an
intensity that varies across the surface of the body according to a
spatially-dependent
albedo $A$. Throughout this paper, we will take $A$ to mean the \emph{spherical}
albedo, the fraction of power incident on a body at a given wavelength
that is scattered back out to space (in all directions).
Note that the spherical albedo
is closely related to the Bond albedo: the Bond albedo is the stellar
flux-weighted integral of $A(\lambda)$ over all wavelengths $\lambda$
\citep[see, e.g.,][]{Seager2010}.

@@

In the limit that the angular size of the star as seen from the planet is
small, the illumination profile on the surface of the planet will
decrease as the cosine of the angle between the surface normal and the star,
reaching zero at the day/night terminator, an angle $\nicefrac{\pi}{2}$
away from the sub-stellar point. However, if the star is sufficiently large
and the planet is sufficiently close-in, rays originating from near the limb
of the star will reach points on the planet surface beyond this angle.
If the stellar radius $R_\star$
is larger than the planet radius $R_\mathrm{p}$, the angular extent of the true
day/night terminator past $\nicefrac{\pi}{2}$ is given by
%
\begin{proof}{tau}
    \label{eq:tau}
    \tau &= \arcsin\left( \frac{1 - \nicefrac{R_\mathrm{p}}{R_\star}}{\nicefrac{a}{R_\star}} \right)
\end{proof}
%
where $a$ the semi-major axis of the orbit (where we implictly assume
the eccentricity is zero).
For planets like the Earth, this quantity is only about $0.26^\circ$,
resulting in a negligible effect on the planet's light curve. However,
for very close-in planets, the effect can be significant
\citep{Knuth2017,Carter2019}. For instance,
the hot Jupiter Kelt-9b has $\nicefrac{R_\mathrm{p}}{R_\star} = 0.083$ and
$\nicefrac{a}{R_\star} = 3.16$
planet \citep{Wong2019}. Assuming zero eccentricity and ignoring any
stellar oblateness
\citep[see][]{Ahlers2020},
the day/night terminator extends $\tau \approx 17^\circ$ past the limb
of that planet. Figure~\ref{fig:extended} shows the illumination profile
of this planet in a Mollweide projection, with the sub-stellar point at
the center, for the point source approximation (left) and accounting for
the finite size of the star (right). In addition to the
displaced day/night terminator, the main difference between
the two profiles is the sub-stellar intensity, which is significantly higher
in the extended source case. This is due to the simple fact that in the point
source case the illumination source is placed at the center of the star,
which is one stellar radius farther from the planet than the point closest
to the planet (the sub-planetary point) in the extended source case.
Once accounting for this difference, the fractional change in the intensity
on the planet away from the sub-stellar point is similar in both cases,
and the intensity anywhere beyond $\nicefrac{\pi}{2}$ is less than one-tenth
the peak value.

@@

We would also like to emphasize that while spherical harmonics are
a convenient basis for the purpose of computing light curves, they
have a significant drawback when it comes to modeling real planetary
surfaces: it can be difficult to strictly enforce physical
values of the albedo everywhere on the surface map when doing inference.
That is because there is no analytic way to determine whether a
spherical harmonic representation is positive-valued
(or restricted to a given range) everywhere on the sphere.
Instead, this must be checked numerically, by evaluating the function on a
discrete grid. This makes it somewhat cumbersome to implement positivity as a prior
when doing inference; in particular, this prior cannot be expressed as
a Gaussian, so the analytic expression for the posterior
discussed in \S\ref{sec:linearity} will generally have nonzero support for
negative albedo values. This is particularly problematic when the data is
not very constraining and the posterior is prior-dominated, as
positivity can be an \emph{extremely} informative prior
\citep[e.g.,][]{Fienup1978}.
We therefore recommend that in such cases
users of the \starry algorithm use HMC/MCMC to do inference, either
(1) sampling over the spherical
harmonic coefficients $\mathbf{y}$ and imposing a uniform prior in the range
$[0, 1]$ on the albedo values $\mathbf{a}$
evaluated on a discrete grid on the sphere or
(2) sampling over the albedo values $\mathbf{a}$ with the same uniform prior,
but using $\mathbf{y}$ to compute the actual light curve model
\citep{Bartolic2021}. In both
cases, there exists a linear operator that transforms between
$\mathbf{y}$ and $\mathbf{a}$:
%
\begin{align}
    \label{eq:Py}
    \mathbf{a} = \mathbf{P} \mathbf{y}
\end{align}
%
and
%
\begin{align}
    \label{eq:PInvy}
    \mathbf{y} = \mathbf{P}^+ \mathbf{a}
    \quad,
\end{align}
%
where
%
\begin{align}
    \label{eq:PInv}
    \mathbf{P}^+ =
    (\mathbf{P}^\top \mathbf{P} + \lambda \boldsymbol{\mathsf{I}})^{-1} \mathbf{P}^\top
\end{align}
%
is the pseudoinverse of $\mathbf{P}$, with (small) regularization parameter
$\lambda$ and where $\boldsymbol{\mathsf{I}}$ is the identity matrix.
%
Each row of the matrix $\mathbf{P}$ is constructed from the value
of each of the spherical harmonics at the corresponding point on the
grid; both $\mathbf{P}$ and its inverse may be pre-computed for
efficiency. In both cases, the grid should be fine enough to ensure
positivity over most of the sphere but not so fine as to throttle the
computation; as a rule of thumb, we find that grids with ${\sim}4$ times
as many pixels
as spherical harmonic coefficients are sufficient.
%
The \href{https://starry.readthedocs.io}{documentation}
includes tutorials on how to implement this in practice.

Finally, while the algorithm presented here can be used to model
planet-planet and planet-moon occultations in reflected light, we only
account for the physical blocking of light rays from the planet by the occultor;
we do not account for the attenuation of the reflected light due to the
\emph{shadow} of the occultor on the planet.
This can be an important effect at certain orbital phases
\citep[e.g.,][]{Cabrera2007}. However, since the shadow of the occultor is not necessarily
circular, it is significantly harder to model in closed form. A proper treatment
of this and other effects, such as occultations and shadows due to rings
\citep[e.g.,][]{Arnold2004}, is deferred to future work.
